Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-01 Manuscript/manuscriptPS.tex
Stylistic changes to Results on pure PS simulations with no salt

\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Markus Miettinen}
\affiliation{Potsdam, Germany}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
\affiliation{Southampton, United Kingdom}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\

@@

    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06 & 250:50 & 11174  & 214  & 298  & 200 & ?  & \cite{POPC5POPS1with1MCaClCHARMM}  \\
    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
  \todo{I think that the peak labeling would be good to show also in (A).}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }
\end{figure}

@@

The headgroup and glycerol backbone of PS lipids show wide variety in the order parameters and structures
between different simulation models (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
as previously observed also for PC lipids \cite{botan15}.
The models perform generally less well for PS lipids than for PC lipids 
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
Therefore, the interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

@@

However, the best performing models,
Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
when comparing the PS results to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibit a more asymmetric
and restricted angle distribution for PS than for PC headgroup in
CHARMM36 simulations in best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
Also, the sampled conformations of glycerol backbone
significantly vary between different simulation models 
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which the PS headgroup.

The suggested characteristic conformations of PS headgroup may be useful when interpreting experiments and
can guide the further force field development. However, more accurate simulations are
required to confirm the suggested conformations because the simulated PS headgroup order parameters
are not within the experimental error bars in any of the tested models.

>>>>

\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Markus Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
\affiliation{Southampton, United Kingdom}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & 11207 & 0  & 298  & 200 & 180   & \cite{POPC5POPS1noCaClCHARMM}  \\

@@

    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  & 1.06 & 250:50  & 11174  & 214  & 298  & 200 & 180  & \cite{POPC5POPS1with1MCaClCHARMM}  \\
    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}        & 0    & 150:150 & 10785    & 0  & 298  & 200 & 180   & \cite{POPC1POPS1noCaClCHARMM}  \\

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
  \todo{I think that the peak labeling would be good to show also in (A).}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp-eps-converted-to.pdf}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }
\end{figure}

@@

The headgroup and glycerol backbone of PS lipids show a wide variety in the order parameters (Fig.~\ref{HGorderParametersPS})
and structures (Fig.~\ref{HGandGLYstructuresPS}) between different simulation models,
as previously observed also for PC lipids \cite{botan15}.
The models perform generally less well for PS than for PC
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
Therefore, the interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

@@

However, if one excludes the glycerol backbone and focuses only on the headgroup ($\alpha$ and $\beta$) order parameters,
the best performing models,
Slipids, CHARMM36 and CHARMM36ua, do reproduce the larger-than-in-PC forking of the $\alpha$-carbon.
Slipids reproduces also the feature that the $\beta$-carbon order parameter
in PS is significantly smaller than in PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of the carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibits a more asymmetric
and restricted angle distribution for PS than for PC headgroup in
CHARMM36 simulations in best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
Also, the sampled conformations of glycerol backbone
significantly vary between different simulation models 
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which the PS headgroup.

The suggested characteristic conformations of the PS headgroup may be useful when interpreting experiments and
can guide further force field development. However, more accurate MD force fields are
required to confirm the suggested conformations because the simulated PS headgroup order parameters
are not within the experimental error in any of the tested models.

### 2019-02-01 Manuscript/manuscriptPSsuppl.tex
Update authorlist
Stylistic changes to Results on pure PS simulations with no salt

\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Markus Miettinen}
\affiliation{Potsdam, Germany}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
\affiliation{Southampton, United Kingdom}

@@

    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todo{Trajectories and further details to be added by J. Madsen}  \\

@@

Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids
with potassium countertions were ran using CHARMM36 force field.
The data are available from Refs. \citenum{??} \todo{J. Madsen, please write simulation details}. \\

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium} \\
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
were simulated with the additional calcium concentrations
\todo{J. Madsen, please write simulation details}. \\

@@

\pagebreak
\section{Calcium density profiles from simulations with POPC:POPS(5:1) mixture}
\begin{figure}[ht]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixtureALL}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
  \todo{Should we include also counterions into the plot?} \\
\end{figure}

>>>>

\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Fernando Favela}
\affiliation{Mexico}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Markus Miettinen}
\affiliation{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
\affiliation{Southampton, United Kingdom}

@@

    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 250:50 & 11207 & 0  & 298  & 200 & 180   & \citenum{POPC5POPS1noCaClCHARMM}  \\

@@

Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids
with potassium countertions were ran using CHARMM36 force field.
The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and  \citenum{POPC1POPS1noCaClCHARMM}
\todo{J. Madsen, please write simulation details}. \\

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium} \\
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
were simulated with the additional calcium concentrations.
The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
\todo{J. Madsen, please write simulation details}. \\

@@

\pagebreak
\section{Calcium density profiles from simulations with POPC:POPS(5:1) mixture}
\begin{figure}[ht]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixtureALL}
    Ca2+ density profiles from simulations.
  }
  \todo{Should we include also counterions into the plot?} \\
\end{figure}
